---
title: "dev-conc-resp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dev-conc-resp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GeoTox)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
library(scales)
```

### Format data

Split the dose response data into a list by chemical casn.

```{r format_data}
ice_conc_resp <- split(as.data.frame(geo_tox_data$ice), ~casn)
```

### 2-parameter Hill fit

Apply the `fit_hill` function to each of the chemicals.

```{r fits}
fits <- lapply(ice_conc_resp, function(df) {fit_hill(df$logc, df$resp)})
```

### Extract fit parameters

Note that estimates of `tp.sd` and `logAC50.sd` that are `NA` will be replaced with the corresponding `tp` and `logAC50` values. This imputation will be indicated by setting `tp.sd.imputed` and `logAC50.sd.imputed` to `TRUE`.

```{r fit_params}
fit_params <- extract_hill_params(fits)
```

#### Plot fit parameters

```{r}
xylim <- range(with(fit_params, c(tp, tp.sd)), na.rm = T)
ggplot(fit_params, aes(tp, tp.sd)) +
  geom_abline(linetype = 3) +
  geom_point(aes(color = tp.sd.imputed), show.legend = FALSE) +
  coord_cartesian(xlim = xylim, ylim = xylim)
```

```{r}
xylim <- range(with(fit_params, c(logAC50, logAC50.sd)), na.rm = T)
ggplot(fit_params, aes(logAC50, logAC50.sd)) +
  geom_abline(linetype = 3) +
  geom_point(aes(color = logAC50.sd.imputed), show.legend = FALSE) +
  coord_cartesian(xlim = xylim, ylim = xylim)
```

### Plot fits

```{r plot_fits}
log10_x <- seq(-3, 3, length.out = 100)

y <- apply(fit_params, 1, function(par) {
  tp      <- as.numeric(par["tp"])
  logAC50 <- as.numeric(par["logAC50"])
  tp / (1 + 10^(logAC50 - log10_x))
})
colnames(y) <- names(ice_conc_resp)

df <- as_tibble(y) %>% mutate(x = 10^log10_x, .before = 1)

ggplot(df %>% pivot_longer(!x), aes(x, value, color = name)) +
  geom_line(show.legend = FALSE) +
  scale_x_log10(labels = scales::label_math(10^.x, format = log10))
```
