---
title: "GeoTox Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GeoTox Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette covers basic use of package functions. Package data, `geo_tox_data`, is used throughout the examples and details on how it was created can be found in the "GeoTox Package Data" vignette.

```{r setup}
library(GeoTox)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
library(scales)
library(sf)
library(ggpubr)

set.seed(1234)

MC_iter <- 50
```

## Fit Hill model

Split dose response data into a list by chemical casn and then apply the `fit_hill` function to each of the chemicals. Note that estimates of `tp.sd` and `logAC50.sd` that are `NA` will be replaced with the corresponding `tp` and `logAC50` values. This imputation will be indicated by setting `tp.sd.imputed` and `logAC50.sd.imputed` to `TRUE`.

```{r fit_params}
data <- split(geo_tox_data$dose_response, ~casn)
hill_2_params <- fit_hill(data)
hill_3_params <- fit_hill(data, fixed_slope = FALSE)
```

### Plot fits

```{r plot_function}
plot_hill <- function(fit_params) {

  log10_x <- seq(-1, 4, length.out = 100)

  y <- apply(fit_params, 1, function(par) {
    tp      <- as.numeric(par["tp"])
    logAC50 <- as.numeric(par["logAC50"])
    slope   <- as.numeric(par["slope"])
    tp / (1 + 10^(slope * (logAC50 - log10_x)))
  })
  colnames(y) <- fit_params$name

  as_tibble(y) %>%
    mutate(conc = 10^log10_x, .before = 1) %>%
    pivot_longer(!conc, values_to = "resp") %>%
    ggplot(aes(conc, resp, color = name)) +
    geom_line() +
    scale_x_log10(labels = scales::label_math(10^.x, format = log10))
}
```

```{r plot_2_params, fig.height = 3, fig.width = 6, fig.align = "center"}
plot_hill(hill_2_params)
```

```{r plot_3_params, fig.height = 3, fig.width = 6, fig.align = "center"}
plot_hill(hill_3_params)
```

## Simulate Data

```{r simulate_data}
# Age
age <- simulate_age(split(geo_tox_data$age, ~FIPS), n = MC_iter)

# Obesity status
obesity <- simulate_obesity(geo_tox_data$obesity, n = MC_iter)

# Inhalation rate
IR <- simulate_inhalation_rate(age)

# External exposure concentration
C_ext <- simulate_exposure(split(geo_tox_data$exposure, ~FIPS), n = MC_iter)

# Sample from pre-simulated steady-state plasma concentration data
C_ss <- sample_Css(geo_tox_data$simulated_css, age, obesity)
```

## Calculations

```{r calculations}
# Internal dose
D_int <- calc_internal_dose(C_ext, IR, scaling = 1 / 1000)

# in vitro concentration
C_invitro <- calc_invitro_concentration(D_int, C_ss)

# Concentration response
resp <- calc_concentration_response(C_invitro, hill_2_params)
```

## Plots

```{r plot_response, fig.height = 3.5, fig.width = 6, fig.align = "center"}
resp_df <- tibble(
  FIPS = names(resp),
  data = resp
) %>%
  unnest(cols = data) %>%
  pivot_longer(-FIPS, names_to = "health_measure") %>%
  mutate(
    health_measure = factor(
      health_measure,
      levels = c("GCA.Eff", "IA.eff", "GCA.HQ.10", "IA.HQ.10")
    )
  ) %>%
  # Compute summary stats
  summarize(
    median = median(value , na.rm = TRUE),
    quantile_05 = quantile(value, 0.05, na.rm = TRUE, names = FALSE),
    quantile_95 = quantile(value, 0.95, na.rm = TRUE, names = FALSE),
    .by = c(FIPS, health_measure)
  ) %>%
  pivot_longer(
    cols = c(median, quantile_05, quantile_95),
    names_to = "stat"
  ) %>%
  # Order stats
  mutate(
    stat = factor(stat, levels = c("quantile_05", "median", "quantile_95"))
  ) %>%
  # Add county boundaries
  left_join(geo_tox_data$boundaries$county, by = join_by(FIPS)) %>%
  st_as_sf()

make_county_heatmap <- function(df, legend_name) {
  ggplot(df, aes(fill = value)) +
    # Plot county data using fill, hide county borders by setting color = NA
    geom_sf(color = NA) +
    # Add state borders
    geom_sf(data = geo_tox_data$boundaries$state, fill = NA, size = 0.15) +
    # Create separate plots for each stat
    facet_wrap(
      ~stat,
      ncol = 3,
      labeller = labeller(
        stat = c(
          "quantile_05" = "5th Percentile",
          "median" = "Median",
          "quantile_95" = "95th Percentile"
        )
      )
    ) +
    # Add fill scale
    scale_fill_viridis_c(
      name = legend_name,
      direction = -1,
      option = "A",
      trans = "sqrt"
    ) +
    # Theme
    theme_bw() +
    theme(
      text = element_text(size = 12),
      legend.text = element_text(size = 8),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )
}

GCA_Eff_plot <- make_county_heatmap(
  resp_df %>% filter(health_measure == "GCA.Eff"),
  paste("Predicted Response", "Log2 Fold Change", "mRNA Expression",sep = "\n")
)

HQ_10_plot <- make_county_heatmap(
  resp_df %>% filter(health_measure == "IA.HQ.10"),
  "Risk Quotient"
)

ggarrange(
  GCA_Eff_plot , HQ_10_plot,
  labels = c( "A", "B"),
  vjust = 1,
  align = "v",
  nrow = 2,
  font.label = list(size = 20, color = "black", face = "bold"),
  common.legend = FALSE
)
```

```{r plot_exposure, fig.height = 5, fig.width = 6, fig.align = "center"}
exposure_df <- geo_tox_data$exposure %>%
  left_join(hill_2_params, by = join_by(casn == name)) %>%
  left_join(geo_tox_data$boundaries$county, by = join_by(FIPS)) %>%
  st_as_sf()

ggplot(exposure_df, aes(fill = norm)) +
  geom_sf(color = NA) +
  facet_wrap(~chemical, ncol = 2) +
  # Recolor subset as light grey
  geom_sf(
    data = exposure_df %>% filter(mean == 0),
    fill = "light grey",
    color = "light grey",
    lwd = 0.01
  ) +
  # State borders
  geom_sf(data = geo_tox_data$boundaries$state, fill = NA, size = 0.15) +
  scale_fill_viridis_c(
    name = "Normalized\nConcentration",
    direction = -1,
    option = "A"
  ) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    strip.text = element_text(size = 6),
    text = element_text(size = 12),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
