---
title: "GeoTox Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GeoTox Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette covers basic use of package functions. Package data, `geo_tox_data`, is used throughout the examples and details on how it was created can be found in the "GeoTox Package Data" vignette.

```{r setup, message=FALSE}
library(GeoTox)
library(dplyr)
library(tidyverse)
# This loads a version of geo_tox_data run with the full KCC assays and 
load("/ddn/gs1/project/GeoToxPackage/manuscript_data/geo_tox_data.rda")
n <- 500 # Sample size

```

> **NOTE:** The sample size here is the size of the simulated population in each region. This is different than the sample size in the "package_data" vignette, which is used to generate C_ss values for each chemical at specified age and weight combinations.

## Analysis of single assay

Create GeoTox object, run simulations and computations

```{r single}
set.seed(2357)
geoTox <- GeoTox() |> 
  # Set region and group boundaries (for plotting)
  set_boundaries(region = geo_tox_data$boundaries$county,
                 group  = geo_tox_data$boundaries$state) |> 
  # Simulate populations for each region
  simulate_population(age           = split(geo_tox_data$age, ~FIPS),
                      obesity       = geo_tox_data$obesity,
                      exposure      = split(geo_tox_data$exposure, ~FIPS),
                      simulated_css = geo_tox_data$simulated_css,
                      n             = n) |> 
  # Estimated Hill parameters
  set_hill_params(geo_tox_data$dose_response |>
                    filter(endp == "TOX21_H2AX_HTRF_CHO_Agonist_ratio") |>
                    fit_hill(chem = "casn")) |>
  # Calculate response
  calculate_response() |>
  # Perform sensitivity analysis
  sensitivity_analysis()

geoTox
```

Plot outputs

```{r, fig.width = 7, fig.height = 3, fig.align = 'center'}
plot(geoTox)
plot(geoTox, type = "hill")
plot(geoTox, type = "sensitivity")
```

## Analysis of multiple assay

Create GeoTox object, run simulations and computations

```{r multi}
set.seed(2357)
geoTox <- GeoTox() |> 
  # Set region and group boundaries (for plotting)
  set_boundaries(region = geo_tox_data$boundaries$county,
                 group  = geo_tox_data$boundaries$state) |> 
  # Simulate populations for each region
  simulate_population(age           = split(geo_tox_data$age, ~FIPS),
                      obesity       = geo_tox_data$obesity,
                      exposure      = split(geo_tox_data$exposure, ~FIPS),
                      simulated_css = geo_tox_data$simulated_css,
                      n             = n) |> 
  # Estimated Hill parameters
  set_hill_params(geo_tox_data$dose_response |>
                    fit_hill(assay = "endp", chem = "casn")) |> 
  # Calculate response
  calculate_response() |>
  # Perform sensitivity analysis
  sensitivity_analysis()

saveRDS(geoTox,"multi-run-0817.rds")

geoTox
```

Plot outputs

```{r, fig.width = 7, fig.height = 3, fig.align = 'center'}
plot(geoTox)
plot(geoTox, assay = "TOX21_H2AX_HTRF_CHO_Agonist_ratio")
plot(geoTox, type = "hill")
plot(geoTox, type = "sensitivity")
plot(geoTox, type = "sensitivity", assay = "TOX21_H2AX_HTRF_CHO_Agonist_ratio")
```

## Exposure Map

The exposure map is the same for both single and multiple assay analyses. The map shows the distribution of chemical exposure across regions for all chemicals, not just those used in a particular analysis.

```{r, fig.width = 7, fig.height = 3, fig.align = 'center'}
plot(geoTox, type = "exposure", ncol = 5)
```

If other facet labels are present they can be specified using the `chem_label` argument.

```{r, fig.width = 7, fig.height = 3, fig.align = 'center'}
plot(geoTox, type = "exposure", chem_label = "chnm", ncol = 5)
```
