---
title: "Replicate Eccles et al. 2023 Analysis for North Carolina"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GeoTox Package Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package includes example data `geo_tox_data`. Below is a description of the data and example code for how it was gathered.

> **NOTE:** FIPS codes can change. Since data is being pulled from various sources, ensure that the FIPS values can be used to connect data across these sources. For example, in 2022 Connecticut began the process of going from 8 legacy counties to 9 planning regions.

```{r setup}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(readr)
library(stringr)
library(readxl)
library(httk)
library(arrow)
suppressPackageStartupMessages(library(sf))

geo_tox_region_4 <- list()
```

## Chemical data

### Exposure data

Download CMAQ modeled exposure data from <a href="https://www.epa.gov/AirToxScreen" target="_blank">AirToxScreen</a>. Results from AirToxScreen 2020 for chemicals in North Carolina counties are used for this analysis.

```{r Exposure Data Download and Process, eval=FALSE}
tf <- "../inst/manuscript_data/Region4_2020ATS_Ambient_Concentrations.parquet"

# Check if tfa and tfb exist
if (file.exists(tf)) {
  nata <- read_parquet(tf, as_data_frame = FALSE)
} else {
  nata_4a <- read_xlsx("../inst/manuscript_data/Region4a_2020ATS_Ambient_Concentrations.xlsx")
  nata_4b <- read_xlsx("../inst/manuscript_data/Region4b_2020ATS_Ambient_Concentrations.xlsx")
  nata <- bind_rows(nata_4a, nata_4b)
  write_parquet(nata, tf)
}

# Get the chemical list of NATA data that also has CASRN - downloaded
# from the CompTox Dashboard
chem_list <- read.csv("../inst/manuscript_data/Chemical List NATADB-2024-06-05.csv")
chem_list$PREFERRED.NAME <- toupper(chem_list$PREFERRED.NAME)
chem_list$IUPAC.NAME <- toupper(chem_list$IUPAC.NAME)
chem_list <- chem_list[,c("PREFERRED.NAME", "IUPAC.NAME", "CASRN")]

# Select the metadata columns and the chemical columns
drop_cols <- colnames(nata[,c(2,3,6)])
meta_cols <- colnames(nata[,c(1,4,5)])
chemical_cols <- colnames(nata[,7:183])

# Find the common chemicals between the chemical list and the NATA data
common_chem <- intersect(chem_list$PREFERRED.NAME, colnames(nata[,7:183]))

# Using the arrow dplyr syntax, let's combine the two datasets
# then make it long format with the chemical names as the column names
# then we will join the chemical names with the CASRN from the chemical list

nata_long <- nata %>%
  select(-all_of(drop_cols)) %>% # Remove the columns that are not needed
  select(all_of(meta_cols), all_of(common_chem)) %>% # Select the metadata and chemical columns
  collect() %>% # Collect the data - pivot_longer requires it be a dataframe and not a parquet promise
  pivot_longer(cols = all_of(common_chem), names_to = "chemical", values_to = "value") %>% # Pivot
  left_join(chem_list, by = c("chemical" = "PREFERRED.NAME"), keep = FALSE) |>  # Join with the chemical
  left_join(chem_list, by = c("chemical" = "IUPAC.NAME"), keep = FALSE) |> # Join with the chemical
  mutate(CASRN = coalesce(CASRN.x, CASRN.y)) %>% # Combine the CASRN columns
  select(-CASRN.x, -CASRN.y) # Remove the CASRN columns

```


```{r, eval=FALSE}

# Normalization function
min_max_norm = function(x) {
  min_x <- min(x, na.rm = TRUE)
  max_x <- max(x, na.rm = TRUE)
  if (min_x == max_x) {
    rep(0, length(x))
  } else {
    (x - min_x) / (max_x - min_x)
  }
}

geo_tox_region_4$exposure <- nata_long %>%
  # Aggregate chemicals by county
  group_by(FIPS, chemical, CASRN) %>%
  # Calculate the mean and sd of every chemical in each county
  summarise(mean = mean(value, na.rm = TRUE), sd = sd(value, na.rm = TRUE)) %>%
  # calculate a normalized value for each chemical in each county 
  ungroup() %>%
  mutate(norm = min_max_norm(mean), .by = chemical) 
```

### Dose-response data


A subset of data for the "LTEA_HepaRG_CYP1A1_up" assay is included in the package data as `geo_tox_data$dose_response`.

```{r, eval=FALSE}

load("../inst/manuscript_data/LTEA_HepaRG_CYP1A1_up 41 chems for Kyle 220131.RData")

geo_tox_region_4$dose_response <- cdat %>%
  as_tibble() %>%
  # Keep "LTEA_HepaRG_CYP1A1_up" assay
  filter(aenm == "LTEA_HepaRG_CYP1A1_up") %>%
  # Keep selected columns
  select(chnm, casn, logc, resp)



load("../inst/mansuscript_data/aeid_851_1000_ICE_cHTS_chems_for_Kyle_231120.RData")

geo_tox_region_4$dose_response <- cdat2_h %>%
  # Keep "LTEA_HepaRG_CYP1A1_up" assay
  filter(aenm == "LTEA_HepaRG_CYP1A1_up") %>%
  # Keep selected columns
  select(chnm, casn, logc, resp)

```

### Subset exposure and dose-response data

Use <a href="https://ice.ntp.niehs.nih.gov/" target="_blank">ICE</a> cHTS data to filter desired chemicals. Retain only those chemicals found in both the exposure and dose-response datasets.

```{r, eval=FALSE}

# Use cHTS data to find active chemicals

casn_hits <- unique(geo_tox_region_4$dose_response$casn)

# Add CASN to exposure, only keep those found in dose-response as well
# Replace with CASN if found for exposure data
geo_tox_region_4$exposure <- geo_tox_region_4$exposure %>%
  # Keep cHTS active chemicals
  filter(CASRN %in% casn_hits)

# Update dose-response to keep only those remaining in exposure
geo_tox_region_4$dose_response <- geo_tox_region_4$dose_response %>%
  filter(casn %in% unique(geo_tox_region_4$exposure$CASRN))
```

## Population data
s
### Age

Download age data from the <a href="https://www.census.gov/" target="_blank">U.S. Census Bureau</a> by searching for "County Population by Characteristics". A subset of data for North Carolina from <a href="https://www.census.gov/data/datasets/time-series/demo/popest/2010s-counties-detail.html" target="_blank">2019</a> is included in the package data as `geo_tox_data$age`.

```{r, eval=FALSE}
# Data for North Carolina
age <- read_csv("cc-est2019-alldata-37.csv")

geo_tox_data$age <- age %>%
  # 7/1/2019 population estimate
  filter(YEAR == 12) %>%
  # Create FIPS
  mutate(FIPS = str_c(STATE, COUNTY)) %>%
  # Keep selected columns
  select(FIPS, AGEGRP, TOT_POP)
```

### Obesity

Follow the "Data Portal" link from <a href="https://www.cdc.gov/places/index.html" target="_blank">CDC PLACES</a> and search for "places county data". Go to the desired dataset webpage, for example <a href="https://data.cdc.gov/500-Cities-Places/PLACES-County-Data-GIS-Friendly-Format-2020-releas/mssc-ksj7/about_data" target="_blank">2020 county data</a>, and download the data by selecting Actions &rarr; API &rarr; Download file. A subset of data for North Carolina is included in the package data as `geo_tox_data$obesity`.

```{r, eval=FALSE}
places <- read_csv("PLACES__County_Data__GIS_Friendly_Format___2020_release.csv")

# Convert confidence interval to standard deviation
extract_SD <- function(x) {
  range <- as.numeric(str_split_1(str_sub(x, 2, -2), ","))
  diff(range) / 3.92
}

geo_tox_data$obesity <- places %>%
  # North Carolina Counties
  filter(StateAbbr == "NC") %>%
  # Select obesity data
  select(FIPS = CountyFIPS, OBESITY_CrudePrev, OBESITY_Crude95CI) %>%
  # Change confidence interval to standard deviation
  rowwise() %>%
  mutate(OBESITY_SD = extract_SD(OBESITY_Crude95CI)) %>%
  ungroup() %>%
  select(-OBESITY_Crude95CI)
```

### Steady-state plasma concentration (Css)

Use `httk` to generate Css values for combinations of age group and weight status for each chemical. The generation of these values is a time-intensive step, so one approach is to generate populations of Css values initially and then sample them later. The number of samples in this example dataset is very small and should be increased if using this method.

```{r, eval=FALSE}
load_sipes2017()

set.seed(2345)
n_samples <- 50

# Define population demographics for httk simulation
# TODO where do these age groups come from?
# They don't match those from the age data source or the table in simulate_inhalation_rate().
# Are they age groups specific to httk?
pop_demo <- cross_join(
  tibble(
    age_group = list(
      c(0, 2), c(3, 5), c(6, 10), c(11, 15), c(16, 20), c(21, 30),
      c(31, 40), c(41, 50), c(51, 60), c(61, 70), c(71, 100)
    )
  ),
  tibble(
    weight = c("Normal", "Obese")
  )
) %>%
  # Create column of lower age_group values
  rowwise() %>%
  mutate(age_min = age_group[1]) %>%
  ungroup()

# Create wrapper function around httk steps
simulate_css <- function(
    chem.cas, agelim_years, weight_category, samples, verbose = TRUE
) {

  if (verbose) {
    cat(
      chem.cas,
      paste0("(", paste(agelim_years, collapse = ", "), ")"),
      weight_category,
      "\n"
    )
  }

  httkpop <- list(
    method = "vi",
    gendernum = NULL,
    agelim_years = agelim_years,
    agelim_months = NULL,
    weight_category = weight_category,
    reths = c(
      "Mexican American",
      "Other Hispanic",
      "Non-Hispanic White",
      "Non-Hispanic Black",
      "Other"
    )
  )

  css <- try(
    suppressWarnings({
      mcs <- create_mc_samples(
        chem.cas = chem.cas,
        samples = samples,
        httkpop.generate.arg.list = httkpop,
        suppress.messages = TRUE
      )

      calc_analytic_css(
        chem.cas = chem.cas,
        parameters = mcs,
        model = "3compartmentss",
        suppress.messages = TRUE
      )
    }),
    silent = TRUE
  )

  # Return
  if (is(css, "try-error")) {
    warning(paste0("simulate_css failed to generate data for CASN ", chem.cas))
    list(NA)
  } else {
    list(css)
  }
}

# Get CASN
casn <- unique(geo_tox_data$dose_response$casn)

# Simulate Css values
simulated_css <- lapply(casn, function(chem.cas) {
  pop_demo %>%
    rowwise() %>%
    mutate(
      css = simulate_css(.env$chem.cas, age_group, weight, .env$n_samples)
    ) %>%
    ungroup()
})
simulated_css <- setNames(simulated_css, casn)

# Get median Css values for each age_group
simulated_css <- lapply(
  simulated_css,
  function(cas_df) {
    cas_df %>%
      nest(.by = age_group) %>%
      mutate(
        age_median_css = sapply(data, function(df) median(unlist(df$css)))
      ) %>%
      unnest(data)
  }
)

# Get median Css values for each weight
simulated_css <- lapply(
  simulated_css,
  function(cas_df) {
    cas_df %>%
      nest(.by = weight) %>%
      mutate(
        weight_median_css = sapply(data, function(df) median(unlist(df$css)))
      ) %>%
      unnest(data) %>%
      arrange(age_min, weight)
  }
)

geo_tox_data$simulated_css <- simulated_css
```

### Subset exposure, dose-response and Css data

Retain only those chemicals found in exposure, dose-response and Css datasets.

```{r, eval=FALSE}
idx <- sapply(
  lapply(geo_tox_data$simulated_css, "[[", "age_median_css"),
  function(x) any(is.na(x))
)

if (any(idx)) {

  idx <- which(!idx)
  casn <- names(geo_tox_data$simulated_css)[idx]

  geo_tox_data$exposure <- geo_tox_data$exposure %>%
    filter(casn %in% .env$casn)

  geo_tox_data$dose_response <- geo_tox_data$dose_response %>%
    filter(casn %in% .env$casn)

  geo_tox_data$simulated_css <- geo_tox_data$simulated_css[idx]

}
```

## County/State boundaries

Download cartographic boundary files for counties and states from the <a href="https://www.census.gov/geographies/mapping-files/time-series/geo/cartographic-boundary.html" target="_blank">U.S. Census Bureau</a>. The geometry data for North Carolina counties and the state are included in the package data as `geo_tox_data$boundaries`.

```{r, eval=FALSE}
county <- st_read("cb_2019_us_county_5m.shp")
state <- st_read("cb_2019_us_state_5m.shp")

geo_tox_data$boundaries <- list(
  county = county %>%
    filter(STATEFP == 37) %>%
    select(FIPS = GEOID, geometry),
  state = state %>%
    filter(STATEFP == 37) %>%
    select(geometry)
)
```
